Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-07-25 Paper/assa.tex
..

\section{Active subspaces and sensitivity analysis}
\label{sec:assa}

\subsection{Active subspaces}
\label{sub:ac}
We recall some background regarding active subspaces~\cite{Constantine:2015}.

\subsection{Global sensitivity measures}
\label{sub:gsa}
Define the total Sobol' indices $T_i(f)$ and discuss their use.

For $f: \Omega \to \R$, we consider the DGSMs~\cite{Sobol:2009},
\[
    \nu_i(f) := \E{\left(\frac{\partial f}{\partial\xi_i}\right)^2} =
                  \int_\Omega 
                  \left(\frac{\partial f}{\partial\xi_i}\right)^2
                  \pi(\vec{\xi})d\vec{\xi}, \quad i = 1, \ldots, \Np.   
\]
Here $\pi$ is the joint PDF of $\xi$. 
Let $\mat{C} = \E{\nabla f \nabla f^T}$ and note that
\[
   \nu_i(f) = \vec{e}_i^T \mat{C} \vec{e}_i, 
\]
where $\vec{e}_i$ is the $i$th coordinate vector in $\R^\Np$.
This also shows that $\sum_{j=1}^\Np \nu_i(f) = \trace(\mat{C})$.
We consider the spectral decomposition of $\mat{C}$, given by  
$\mat{C} = \sum_{k=1}^\Np \lambda_k \vec{u}_k \vec{u}_k^T$, where 
$\lambda_i$ are the (non-negative) eigenvalues of $\mat{C}$ and
$\vec{u}_k$ are the corresponding (orthonormal) eigenvectors.
We note that,
\begin{equation}\label{equ:spectral_DGSM}
\nu_i(f) = \vec{e}_i^T \Big(\sum_{k=1}^\Np \lambda_k \vec{u}_k \vec{u}_k^T\Big) \vec{e}_i
 = \sum_{k=1}^\Np \lambda_k \ip{\vec{e}_i}{\vec{u}_k}^2, 
\end{equation}
which gives a spectral representation for the DGSMs. Truncating the summation 
leads to the notion of activity scores introduced in~\cite{Diaz:2016,Constantine:2017}
defined as follows: 
\[
   \alpha_i(f; r) =  \sum_{k=1}^r \lambda_k \ip{\vec{e}_i}{\vec{u}_k}^2,
   \quad j = 1, \ldots, \Np, \quad r \leq \Np.
\]
The activity scores connect ideas from active subspaces and global sensitivity
analysis, and can be used to approximate DGSMs.  The following result, which
can be found in~\cite{Diaz:2016,Constantine:2017}, quantifies the error in this
approximation. We provide a short proof for completeness. 
\begin{proposition}\label{prp:dgsm_bound} 
For $1 \leq r \leq \Np$,
\[
0 \leq \nu_i(f) - \alpha_i(f; r) \leq \lambda_{r+1}, \quad i = 1, \ldots, \Np.
\] 
\end{proposition}
\begin{proof}

>>>>

\section{Active subspaces and sensitivity analysis}
\label{sec:assa}
In this section, we provide key elements of the active subspace methodology in~\ref{sub:ac}, and
provide theoretical connections between the global sensitivity measures: DGSMs, activity scores,
and the total Sobol index in~\ref{sub:gsa}.

\subsection{Active subspaces}
\label{sub:ac}
%We recall some background regarding active subspaces~\cite{Constantine:2015}.
As mentioned earlier, the active subspace is low-dimensional linear subspace that constitutes
important directions in a model's input space~\cite{Constantine:2015}. The effective variability
in the model output due to uncertain inputs is predominantly captured along these directions.
Mathematically, these directions are essentially the eigenvectors of a positive semi-definite
matrix, $\mat{C}$, defined as follows:
%
\be
\mat{C} = \int_\Omega (\nabla_{\bm{\xi}}f)(\nabla_{\bm{\xi}}f)^\top dP_{\bm{\xi}}
\label{eq:C}
\ee
%
where $\Omega$:~$[-1,1]^{N_p}$ is the domain of $\bm{\xi}$, and $N_p$ is the number of uncertain
inputs. $dP_{\bm{\xi}}$ = $p(\bm{\xi})d\bm{\xi}$, where $p(\bm{\xi})$ is the joint probability
distribution of $\bm{\xi}$. Note that the function, $f$ is considered to be continuous and differentiable
in the considered parameter domain. Hence, it is possible that a given $f(\bm{\xi})$ might not admit an
active subspace. However, it is of remarkable interest to investigate if one exists since the subspace could
be exploited to reduce the dimensionality of the problem and hence the associated computational effort.
$\nabla_{\bm{\xi}}f$ denotes a vector with individual components
being partial derivatives of $f$ with respect to the $i^\text{th}$ input, $\xi_i$. Since $\mat{C}$ is symmetric and
positive semi-definite, it admits a real eigenvalue decomposition:
%
\be
\mat{C} = \bm{W}\bm{\Lambda}\bm{W}^\top
\ee
%
where $\bm{\Lambda}$ = diag($\lambda_1,\ldots,\lambda_{N_p}$) with $\lambda_i$'s
 sorted in descending order, $i:\{1,2,\ldots,N_p\}$. The eigenpairs are partitioned about
 the $i^{\text{th}}$ eigenvalue such that the ratio, $\left(\frac{\lambda_i}{\lambda_{i+1}}\right)$ $\gg$ 1
 as follows:
 %
 \be
 \bm{W} = [\bm{W}_1~\bm{W}_2],~~\bm{\Lambda} = \begin{bmatrix}\bm{\Lambda}_1 & \\  &
  \bm{\Lambda}_2 \\
\end{bmatrix}
 \ee
 %
 where $\bm{W}_1$ comprises the dominant eigenspace of $\mat{C}$ regarded as the active subspace,
 and $\bm{\Lambda}_1$ is the corresponding set of eigenvalues. The active subspace is used to 
 construct the so-called active variables, $\bm{w}_1^\top \bm{\xi}$ and $f(\bm{\xi}(\theta))$ is transformed
 as $G(\bm{w}_1^\top \bm{\xi})$. In practice, the set of samples, $\bm{\xi}_i$'s are projected to the
 physical space as $\bm{\theta}_i$'s at which the model output, $f(\bm{\xi}(\theta))$ is computed. 
 The matrix, $\mat{C}$ is typically approximated using pseudo-random sampling techniques such 
 as Monte Carlo and Latin hypercube sampling. The integral in~\eqref{eq:C} is replaced with a
 summation as follows:
 %
 \be
 \mat{C}\approx \hat{\mat{C}} = \frac{1}{N}\sum\limits_{i=1}^{N} 
 (\nabla_{\bm{\xi}}f(\bm{\xi}_i))(\nabla_{\bm{\xi}}f(\bm{\xi}_i))^\top
 = \hat{\bm{W}}\hat{\bm{\Lambda}}\hat{\bm{W}}^\top
 \ee
 %
 It has been shown that the accuracy of approximated dominant eigenspace, $\hat{\bm{W}}_1$ is inversely
 proportional to the difference between the smallest eigenvalue in $\hat{\bm{\Lambda}}_1$ and the
 largest eigenvalue in $\hat{\bm{\Lambda}}_2$~\cite{Constantine:2014}.

 Components of the eigenvectors
 in the active subspace could be used for estimating the so-called activity scores as a measure for
 global sensitivity and also be used for approximating the DGSMs as discussed in the following section.

\subsection{Global sensitivity measures}
\label{sub:gsa}
The total Sobol' index ($T_i(f)$) of a model output, $f(\theta_i)$ quantifies the total contribution of the
input, $\theta_i$ to the variance of the output. Mathematically, this can be expressed as follows:
%
\be
T_i(f) = 1 - 
\frac{\V[\mathbb{E}(f|\bm{\theta}_{\sim i})]}{\V(f)},
\label{eq:total}
\ee
%
where $\bm{\theta}_{\sim i}$ is a vector of all but $i^\text{th}$ uncertain input, and $\V$ denotes the variance.
Determining converged estimates of $T_i(f)$ typically involves tens of thousands of model runs and is therefore
prohibitive in the case of compute-intensive applications. The DGSMs~\cite{Sobol:2009}
provide a means for approximating the upper bound on $T_i(f)$ with fewer computations as 
discussed in~\cite{Vohra:2018}. 
\smallskip

For $f: \Omega \to \R$, we consider the DGSMs,
\[
    \nu_i(f) := \E{\left(\frac{\partial f}{\partial\xi_i}\right)^2} =
                  \int_\Omega 
                  \left(\frac{\partial f}{\partial\xi_i}\right)^2
                  \pi(\vec{\xi})d\vec{\xi}, \quad i = 1, \ldots, \Np.   
\]
Here $\pi$ is the joint PDF of $\xi$. 
Let $\mat{C} = \E{\nabla f \nabla f^\top}$ and note that
\[
   \nu_i(f) = \vec{e}_i^\top \mat{C} \vec{e}_i, 
\]
where $\vec{e}_i$ is the $i$th coordinate vector in $\R^\Np$.
This also shows that $\sum_{j=1}^\Np \nu_i(f) = \trace(\mat{C})$.
We consider the spectral decomposition of $\mat{C}$, given by  
$\mat{C} = \sum_{k=1}^\Np \lambda_k \vec{u}_k \vec{u}_k^\top$, where 
$\lambda_i$ are the (non-negative) eigenvalues of $\mat{C}$ and
$\vec{u}_k$ are the corresponding (orthonormal) eigenvectors.
We note that,
\begin{equation}\label{equ:spectral_DGSM}
\nu_i(f) = \vec{e}_i^T \Big(\sum_{k=1}^\Np \lambda_k \vec{u}_k \vec{u}_k^\top\Big) \vec{e}_i
 = \sum_{k=1}^\Np \lambda_k \ip{\vec{e}_i}{\vec{u}_k}^2, 
\end{equation}
which gives a spectral representation for the DGSMs. Truncating the summation 
leads to the notion of activity scores introduced in~\cite{Diaz:2016,Constantine:2017}
defined as follows: 
\[
   \alpha_i(f; r) =  \sum_{k=1}^r \lambda_k \ip{\vec{e}_i}{\vec{u}_k}^2,
   \quad j = 1, \ldots, \Np, \quad r \leq \Np.
\]
The activity scores connect ideas from active subspaces and global sensitivity
analysis, and can be used to approximate DGSMs.  The following result, which
can be found in~\cite{Diaz:2016,Constantine:2017}, quantifies the error in this
approximation. We provide a short proof for completeness. 
\begin{proposition}\label{prp:dgsm_bound} 
For $1 \leq r \leq \Np$,
\[
0 \leq \nu_i(f) - \alpha_i(f; r) \leq \lambda_{r+1}, \quad i = 1, \ldots, \Np.
\] 
\end{proposition}
\begin{proof}

### 2018-07-25 Paper/intro.tex
..

We are interested in investigating the impact of uncertainty in the rate-controlling parameters, 
pre-exponent~($A_i$) and the activation energy~($E_{a,i}$) on ignition delay. The $A_i$'s are
considered to be uniformly distributed in the interval: $[0.9A_i^\ast, 1.1A_i^\ast]$.
The $E_{a,i}$'s for all reactions except $\mathcal{R}_6$ -- $\mathcal{R}_9$ and $\mathcal{R}_{13}$
are considered to be uncertain and uniformly distributed in the interval: $[0.99E_{a,i}^\ast, 1.01E_{a,i}^\ast]$.
$A_i^\ast$ and $E_{a,i}^\ast$ are nominal estimates corresponding to the $i^\text{th}$ reaction
as provided in~\cite{Yetter:1991}. Hence, the total number of uncertain parameters is 33 which makes the
present problem reasonably high-dimensional involving a significantly large amount of computational
effort. To reduce the effort, we explore two strategies: one involving the use of gradients (grad-based)
and the other involving local linear approximation which does require gradient computation (grad-free).
These strategies are based on algorithms 1.1 and 1.2 provided  in~\cite{Constantine:2015}. The amount of 
computational effort associated with the grad-free approach is significantly smaller than the grad-based
approach. In section~\ref{sec:method}, we implement the two strategies to a simpler problem accounting for the
uncertainty only in the pre-exponents ($A_i$'s) of the rate constants in the H$_2$/O$_2$ reaction mechanism.
The two strategies are observed to yield consistent results which encourages the use of the grad-free approach
for discovering the active subspace especially in high-dimensional problems. In section~\ref{sec:app}, we present
our findings based on the implementation of the grad-free approach to the original 33-dimensional application.

@@

\paragraph{Contributions}
Main contributions of this paper are as follows: (1) A perspective on activity scores as means for approximating
the DGSMs is provided. (2) A general relationship between DGSMs, activity scores, and the total Sobol' index 
is established. (3) Iterative gradient-based and gradient-free approaches for computing the active subspace are
 demonstrated. The two approaches are shown to be consistent and yield computational gains compared to the 
 conventional approach. As mentioned above, the gains are greater in the case of grad-free approach. (4) The
 grad-free approach is implemented for discovering an active subspace in a 33-dimensional parameter space 
 for the H$_2$/O$_2$ reaction kinetics application.

>>>>

We are interested in investigating the impact of uncertainty in the rate-controlling parameters, 
pre-exponent~($A_i$) and the activation energy~($E_{a,i}$) on ignition delay. The $A_i$'s are
considered to be uniformly distributed in the interval: $[0.9A_i^\ast, 1.1A_i^\ast]$.
The $E_{a,i}$'s for all reactions except $\mathcal{R}_6$ -- $\mathcal{R}_9$ and $\mathcal{R}_{13}$
are considered to be uncertain and uniformly distributed in the interval: $[0.99E_{a,i}^\ast, 1.01E_{a,i}^\ast]$.
$A_i^\ast$ and $E_{a,i}^\ast$ are nominal estimates corresponding to the $i^\text{th}$ reaction
as provided in~\cite{Yetter:1991}. Hence, the total number of uncertain parameters is 33 which makes the
present problem reasonably high-dimensional involving a significantly large amount of computational
effort. To reduce the effort, we explore two strategies: one involving the use of gradients (grad-based)
and the other involving local linear approximation which does require gradient computation (grad-free).
These strategies are based on algorithms 1.1 and 1.2 provided  in~\cite{Constantine:2015}. The amount of 
computational effort associated with the grad-free approach is significantly smaller than the grad-based
approach as discussed later in section~\ref{sec:method}.

@@

\paragraph{Contributions}
Main contributions of this paper are as follows: (1) A perspective on activity scores as means for approximating
the DGSMs is provided. (2) A general relationship between DGSMs, activity scores, and the total Sobol' index 
is established. (3) Iterative gradient-based and gradient-free approaches for computing the active subspace are
 demonstrated. The two approaches are shown to be consistent and yield greater
computational gains compared to the 
 conventional approach. As mentioned above, the gains are greater in the case of grad-free approach. (4) The
 grad-free approach is implemented for discovering an active subspace in a 33-dimensional parameter space 
 for the H$_2$/O$_2$ reaction kinetics application.
